---
title: "5.1 Data Frames"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = TRUE)
```

---

## 5.1.1: Introduction to Data Frames

---

### 2-Dimensional Data with Multiple Types

Previously we learned how to store 2D data in a matrix. However, in biology we often have 2D data with multiple data types. Let's look at a real example.

“The IMPC is an international effort by 21 research institutions to identify the function of every protein-coding gene in the mouse genome.” (https://www.mousephenotype.org) To achieve this the IMPC knocks out each mouse gene individually and uses a panel of hundreds of standardized tests to assess mouse phenotypes. This includes (but is not limited to): assessment of behavioural tendencies, x-rays to examine the skeletal system, blood tests, and observations of morphological differences. These can be compared to controls to determine which phenotypes are affected by the knocked out gene.

For example, here are data for 4 mice with the Ap4e1 gene knocked out (mutant mice), and data for 4 control mice. This table shows the results of a test to assess the hemoglobin content (in g/dL) along with mouse sex and mutant status. (Data from: https://www.mousephenotype.org)

| Sex    | Mutant | Hemoglobin (g/dL) |
| ------ | ------ | ----------------- |
| female | TRUE   | 14.9              |
| female | TRUE   | 15.1              |
| male   | TRUE   | 14.6              |
| male   | TRUE   | 15.2              |
| female | FALSE  | 13.4              |
| female | FALSE  | 13.9              |
| male   | FALSE  | 13.4              |
| male   | FALSE  | 14.9              |

---

### Data Frames

We cannot store this data in a matrix as sex should be stored as character type, mutant status as logical type, and hemoglobin content as numeric type. However, we can store the data in a *data frame*.

A data frame is an R data structure that stores a collection of vectors. Each vector can be a different data type, but all vectors must be the same length. A data frame arranges these vectors into columns.

Data frames are like spreadsheets. In a data frame, generally, each row represents a sample (individual patients, mice, genes, tissues, proteins, tubes, etc.) and each column represents a feature or attribute of the samples. These are measurements, categories, and/or information about each sample.

For our Ap4e1 data, we have 8 individual samples, two categorical values (sex and mutant status) and one measurement (hemoglobin content).

---

## 5.1.2: Data Frame Creation

---

### The `data.frame()` Function

To create a data frame, we can use the `data.frame()` function. The general syntax is:

    data.frame(column_name1=vector1, column_name2=vector2, column_name3=vector3)
    
As mentioned earlier, a data frame is a collection of vectors of different types, arranged into columns. We therefore need to provide each vector along with a column name to the function. All of the vectors should be of the same length! If the vectors are not of the same length, R will repeat the shorter vectors until they are the same length as the longest vector.

Let's create a data frame for our Ap4e1 mutant mice:

| Sex    | Mutant | Hemoglobin (g/dL) |
| ------ | ------ | ----------------- |
| female | TRUE   | 14.9              |
| female | TRUE   | 15.1              |
| male   | TRUE   | 14.6              |
| male   | TRUE   | 15.2              |

It will have three vectors - "sex", "mutant", and "HG" - all of length 4. Note that in order to interpret rows as individual samples, we must include the data in the vectors such that the values at each index correspond to the same sample. For example here, if we look at the value at the third index in each vector, they all correspond to the same mouse:

```{r}
Ap4e1_HG_mutant <- data.frame(sex=c("female", "female", "male", "male"), 
                              mutant=c(TRUE, TRUE, TRUE, TRUE), 
                              HG=c(14.9, 15.1, 14.6, 15.2))
Ap4e1_HG_mutant
```

Looking at the output of a data frame we can see that it looks quite different from a matrix. In R markdown, data frames are displayed in a table format to make them easier to read. The data type of each column is displayed below the column name: "chr" for character, "lgl" for logical, and "dbl" for double. 

### Row & Column Names

The data frame output also shows us how many rows there are at the bottom, but note that row indices or names are not shown. Unlike matrices, data frames must have row names. R automatically assigns row names to data frames on creation. The default row names will always be the positional index of the row, but as a character value (because it is a name):

```{r}
rownames(Ap4e1_HG_mutant)
```

You can see that in `Ap4e1_HG_mutant` the row names are simply the row indices. You can assign row names to a data frame with two methods:

- During data frame creation: `data.frame(..., row.names=c(...))`
- After data frame creation: `rownames(dataframe) <- c(...)`

Row names will be displayed in data frame output if you assign the row names to something other than the default:

```{r}
rownames(Ap4e1_HG_mutant) <- c("m1", "m2", "m3", "m4")
Ap4e1_HG_mutant
```

Row names can be reverted to the original automatic row names by setting them to `NULL`:

    rownames(data.frame) <- NULL

```{r}
rownames(Ap4e1_HG_mutant) <- NULL
rownames(Ap4e1_HG_mutant)
```

Similarly, data frames must have column names. If you create a data frame without assigning column names, R will automatically name your columns. These names are long and difficult to work with as they are representations of the entire column. 

Let's see what happens if we create a data frame for the control mice, without providing column names to the `data.frame()` function:

| Sex    | Mutant | Hemoglobin (g/dL) |
| ------ | ------ | ----------------- |
| female | FALSE  | 13.4              |
| female | FALSE  | 13.9              |
| male   | FALSE  | 13.4              |
| male   | FALSE  | 14.9              |

```{r}
Ap4e1_HG_control <- 
data.frame(c("female", "female", "male", "male"),
           c(FALSE, FALSE, FALSE, FALSE), 
           c(13.4, 13.9, 13.4, 14.9))
Ap4e1_HG_control
```

As you can see it named the columns as character representations of the column vector values. These names would be very difficult to work with, especially if our data frame had even more rows, leading to even longer column names.

Depending on the size of your screen, you may not be able to see the third column in this data frame. Because these column names are so long, R markdown may not be able to fit all three columns into the display. When this happens, there will be an arrow to the right of the column names, which you can click to view the next column. You will also notice that at the bottom of the data frame, where it states the number of rows, you can see how many columns there are and which columns you are currently viewing.

Column names can be modified with the `colnames()` function, in the same way we modify matrix column names:

    colnames(data_frame) <- c(...)
    
Let's fix the column names in the `Ap4e1_HG_control` data frame:

```{r}
colnames(Ap4e1_HG_control) <- c("sex", "mutant", "HG")
Ap4e1_HG_control
```

---

### Other Data Frame Creation Methods

Data frames can also be created by combining two or more existing data frames. Two data frames with the **same columns** in the **same order** can be combined with the `rbind()` function.

Let's combine our data frames with the mutant and control mice into one:

```{r}
Ap4e1_HG <- rbind(Ap4e1_HG_mutant, Ap4e1_HG_control)
Ap4e1_HG
```

Matrices can also be converted to data frames, which is helpful if you want to add new columns to a matrix that are of a different data type.

For example, let's say we have a matrix with numeric data about a set of cancer patients (stage of cancer and months of treatment):

```{r}
cancer_pts <- matrix(c(2,4,3,8,3,11), nrow=3, dimnames=list(c("pt_A", "pt_B", "pt_C"), c("stage", "months_of_treatment")))
cancer_pts
```

We can convert the matrix into a data frame using the function `as.data.frame()`:

```{r}
cancer_pts_df <- as.data.frame(cancer_pts)
cancer_pts_df
```

We can now use `cbind()` to add a new column to the data frame with character values indicating the type of cancer of each patient, which we wouldn't be able to do if we left the patient data as a matrix due to the different data types:

```{r}
cancer_pts_df <- cbind(cancer_pts_df, cancer_type = c("breast", "pancreatic", "gastric"))
cancer_pts_df
```

---

## 5.1.3: Data Frame Indexing

---

### Positional and Named Indexing

Positional indexing of data frames is the same as it is for matrices:

    data_frame[row_number(s), column_number(s)]

For example, if we want to retrieve the values in the 3rd column for rows one through four:

```{r}
Ap4e1_HG[1:4, 3]
```

Or if we want to get the first, second, fifth, and sixth rows:

```{r}
Ap4e1_HG[c(1,2,5,6), ]
```

The same is true for named indexing:

    data_frame[row_name(s), column_name(s)]

For example, if we want to retrieve sex and mutant status for rows one and seven:

```{r}
Ap4e1_HG[c("1","7"),c("sex", "mutant")]
```

Remember that the row names were automatically assigned when we created this data frame, so in this example the row names are the same as the positional indices. Unless row names are assigned to be meaningful character values, it is generally easier to use positional indices for row indexing:

```{r}
Ap4e1_HG[c(1,7),c("sex", "mutant")]
```

---

### Data Frame Column Indexing

Data frames are actually a type of list, with each column being a different element. They differ from regular lists in that all elements must be vectors of equal length, and in that they are strictly 2-dimensional like matrices.

Because data frames are a type of list we can access columns with the indexing strategies used for lists:

    data_frame$column_name
    data_frame[["column_name"]]

For example, we can retrieve the `mutant` column with the `$` indexing:

```{r}
Ap4e1_HG$mutant
```

Or the `HG` column with `[[]]` indexing:

```{r}
Ap4e1_HG[["HG"]]
```

But we cannot retrieve rows with these methods:

```{r}
Ap4e1_HG[["7"]]
```

Remember that to use indexing with a `$`, column names should never have spaces or start with a number.

With all of these methods there are five unique ways to access a column. For example, to access the fourth element of the third column ("HG"):

```{r}
Ap4e1_HG[4,3]
```

```{r}
Ap4e1_HG[4,"HG"]
```

```{r}
Ap4e1_HG$HG[4]
```

```{r}
Ap4e1_HG[[3]][4]
```

```{r}
Ap4e1_HG[["HG"]][4]
```

When writing your own code, choose the style that is easiest and most convenient for you.

---

## 5.1.4: Data Frame Modification

---

### Modifying Existing Elements

Some matrix modification methods can also be used for data frames:

- Changing a single element

    data_frame[row_index, column_index] <- value

- Changing an entire column

    data_frame[, column_index] <- vector_name

- Changing the column names

    colnames(data_frame) <- vector_name

- Changing the row names

    rownames(data_frame) <- vector_name

---

### Modifying & Adding Rows

Modifying or adding a row is more difficult with data frames than it is with matrices. Because data frame columns can contain different data types, you cannot simply create a vector to replace a row. 

Instead, we can create a data frame with one row that has the same columns as the data frame in which we need to add or replace a row.

        new_row <- data.frame(column_name1 = value1, column_name2 = value2, ...)

To modify a row, assign the new row to the data frame at a specific index.
        
        data_frame[row_index,] <- new_row
        
To add a row, use `rbind()` to bind the row to the end of the data frame.
        
        data_frame <- rbind(old_data_frame, new_row)

Let's remove the last row of the data frame and then add it back:

```{r}
Ap4e1_HG <- Ap4e1_HG[1:7, ]
```

We can add it back by creating a single row data frame and using `rbind()`:

```{r}
mouse8 <- data.frame(sex="male", mutant=FALSE, HG=14.9)
Ap4e1_HG <- rbind(Ap4e1_HG, mouse8)
Ap4e1_HG
```

---

### Modifying & Adding Columns

Matrix column addition methods also work for data frames:

        data_frame[, "column_name"] <- new_vector
        data_frame <- cbind(data_frame, column_name = new_vector)
        
However, since data frame columns have more methods for column indexing, more methods exist for column modifications and additions. We can create new columns (or modify existing ones) with list indexing methods as well:

        data_frame$column_name <- new_vector
        data_frame[["column_name"]] <- new_vector

For example, we have more information about this set of mice, including the weight of each mouse in grams. We can add a vector of weights to `Ap4e1_HG` by creating a new column with `$` indexing:

```{r}
Ap4e1_HG$weight <- c(22.0, 23.1, 29.2, 29.1, 24.3, 23.7, 29.5, 30.9)
```

Another measurement taken for each mouse is grip strength. Grip strength measures the force of strength exerted by each mouse in grams. We can add a new vector with mouse grip strength measurements to `Ap4e1_HG` with `[[]]` indexing: 

```{r}
Ap4e1_HG[["grip"]] <- c(124.08, 176.92, 164.09, 145.33, 235.70, 241.73, 231.11, 210.68)
```

We can see both of the new columns in the output data frame:

```{r}
Ap4e1_HG
```

To properly compare mouse grip strength between mutant and control mice, the measurements must be normalized by the weight of each mouse, as heavier mice will be able to exert greater force due to their higher body weight. To do this we can divide the grip strength column by the weight column, round it and add it as a new column called "grip_norm":

```{r}
Ap4e1_HG$grip_norm <- round(Ap4e1_HG$grip/Ap4e1_HG$weight, 2)
Ap4e1_HG
```

---